Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-06-08 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
In this work we introduce the application of convolutional autoencoder neural networks to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on ${}^{46}$Ar. 
The data we analyze were recorded by an active target time-projection chamber (AT-TPC). Machine learning presents an interesting avenue for researchers operating an AT-TPC, as traditional analysis methods of AT-TPC data are both computationally expensive and fit all particle tracks against the event type of interest. The latter presents a considerable challenge when the space of reactions is not known prior to the analysis.

We explore the performance of the autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well. On simulated data these classifiers achieve an $f1$ score \cite{Chinchor1992} of $f1>0.95$. The VGG16 latent classifier achieves this result with as few as $N=100$ samples, as does the convolutional autoencoder when trained on the VGG16 representations of the particle tracks. On real data, pre-processed with noise filtering, the same models achieve an $f1>0.7$. For unfiltered real data the models achieve an $f1$ value larger than $0.6$. Both of the previous results were found with the classifiers trained on $N=100$ samples. Furthermore, we found that the autoencoder model reduces the variability in the identification of proton events by $64\%$ from the benchmark logistic regression classifier trained on the VGG16 latent space on real experimental data.

On the clustering task, we found that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters, with an adjusted rand index \cite{Hubert1985} ($ARI$) $ > 0.8$.  Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of neural networks to clustering by implementing a mixture of autoencoders algorithm. With this model we improved clustering performance on the real experimental data from an $ARI = 0.17$ to an $ARI = 0.40$. However, the neural network clustering suffers from stability issues necessitating further investigations into this approach.

@@

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{Beceiro2018}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent of bubble chamber times and on par with event classification challenges in particle physics today. {\color{blue} cite something HEP here} More often than not, the reaction channel of interest has one of the lowest cross sections, therefore the physicist is faced with the task of sorting out the corresponding events from the ``background" of other reaction channels.

% motivating ML work:
Because TPCs produce 3-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (akin to analysis in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently. {\color{blue} short discussion on prior work in nuclear/HEP/neutrino} In addition, some ML algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled dataset is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know a priori the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated dataset. In the latter case, there must also exist a validation dataset of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

We begin by exploring a naive K-means approach on the latent space of a pre-trained network. Subsequently, we investigate other clustering methods and the autoencoder based clustering algorithm outlined in section \ref{sec:mixae}.

\subsection{Pre-trained model with K-means}
Building on the work by \citet{Kuchera2019}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical K-means approach to clustering. However, the output from the VGG16 network is very high dimensional which traditionally conflicts with good K-means results. With output vectors in $\R^{8192}$ the ratio between $L_p$ distances goes to one with increasing dimensionality, as shown by \citet{Aggarwal}. However, one of the central assumptions in the authors' finding is that the elements are uniformly distributed in the latent space. They note that it is then possible that if the class information lies in some sub-space of the latent data, $L_p$-norms may provide useful clusterings even in high dimensional spaces. To investigate this, we perform a $L_2$-norm based K-means clustering using the full VGG16 representation and the $10^2$ first principal components only.

>>>>

\begin{abstract}
This article presents novel applications of machine learning methods to the problem of event classification in an active target detector, the Active Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant.

The application of various machine learning algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, we find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of neural networks to clustering by implementing a mixture of autoencoders algorithm, with improved results on real data. However, the neural network clustering suffers from stability issues necessitating further investigations into this approach.

%In this work we introduce the application of convolutional autoencoder neural networks to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on ${}^{46}$Ar. 
%The data we analyze were recorded by an active target time-projection chamber (AT-TPC). Machine learning presents an interesting avenue for researchers operating an AT-TPC, as traditional analysis methods of AT-TPC data are both computationally expensive and fit all particle tracks against the event type of interest. The latter presents a considerable challenge when the space of reactions is not known prior to the analysis.

%We explore the performance of the autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well. On simulated data these classifiers achieve an $f1$ score \cite{Chinchor1992} of $f1>0.95$. The VGG16 latent classifier achieves this result with as few as $N=100$ samples, as does the convolutional autoencoder when trained on the VGG16 representations of the particle tracks. On real data, pre-processed with noise filtering, the same models achieve an $f1>0.7$. For unfiltered real data the models achieve an $f1$ value larger than $0.6$. Both of the previous results were found with the classifiers trained on $N=100$ samples. Furthermore, we found that the autoencoder model reduces the variability in the identification of proton events by $64\%$ from the benchmark logistic regression classifier trained on the VGG16 latent space on real experimental data.

%On the clustering task, we found that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters, with an adjusted rand index \cite{Hubert1985} ($ARI$) $ > 0.8$.  Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of neural networks to clustering by implementing a mixture of autoencoders algorithm. With this model we improved clustering performance on the real experimental data from an $ARI = 0.17$ to an $ARI = 0.40$. However, the neural network clustering suffers from stability issues necessitating further investigations into this approach.

@@

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{BECEIRONOVO2015}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent of bubble chamber times and on par with event classification challenges in particle physics today, see for example the recent review of Mehta {\em et al.} \cite{mehta2019}. More often than not, the reaction channel of interest has one of the lowest cross sections, therefore the physicist is faced with the task of sorting out the corresponding events from the ``background" of other reaction channels.

% motivating ML work:
Because TPCs produce 3-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (akin to analysis in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, some ML algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled dataset is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know a priori the types and behaviors of the reactions present in the detector in order to generate a labeled, simulated dataset. In the latter case, there must also exist a validation dataset of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

We begin by exploring a naive K-means approach on the latent space of a pre-trained network. Subsequently, we investigate other clustering methods and the autoencoder based clustering algorithm outlined in section \ref{sec:mixae}.
{\color{blue} Move methods to pervious section -MPK}
\subsection{Pre-trained model with K-means}
Building on the work by \citet{Kuchera2019}, we also use the VGG16 pre-trained network as a baseline for the clustering performance. We begin by considering a classical K-means approach to clustering. However, the output from the VGG16 network is very high dimensional which traditionally conflicts with good K-means results. With output vectors in $\R^{8192}$ the ratio between $L_p$ distances goes to one with increasing dimensionality, as shown by \citet{Aggarwal}. However, one of the central assumptions in the authors' finding is that the elements are uniformly distributed in the latent space. They note that it is then possible that if the class information lies in some sub-space of the latent data, $L_p$-norms may provide useful clusterings even in high dimensional spaces. To investigate this, we perform a $L_2$-norm based K-means clustering using the full VGG16 representation and the $10^2$ first principal components only.
